The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-26 conf_article/neon.tex
added affilation for me

\author{Alexey Kravatskiy \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{kravtskii.aiu@phystech.edu} \\
\And
Ivan Kozyrev \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{kozyrev.in@phystech.edu} \\
\And
Nikolai Kozlov \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{kozlov.na@phystech.edu} \\
\And
Alexander Vinogradov \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{vinogradov.am@phystech.edu} \\
\And
Daniil Merkulov \\
Moscow Institute of Physics and Technology (MIPT), Skoltech, HSE, AI4Science \\
\texttt{daniil.merkulov@phystech.edu} \\
\And
Ivan Oseledets \\
AIRI, Skoltech \\
\texttt{i.oseledets@skoltech.ru}
}

@@

\begin{abstract}
	In this article, we explore the use of various matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan $k$-norms to introduce a family of Muon-like algorithms we name {\it Fanions}, which happen to be similar to Dion. Then working with the duals of convex combinations of the Ky Fan $k$-norms and the Frobenius norm or the $l_\infty$ norm, we construct the families of {\it F-Fanions} and {\it S-Fanions} respectively. Their most prominent members are {\it F-Muon} and {\it S-Muon}. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that F-Muon and S-Muon are always on par with Muon and S-Muo, while on fine-tuning of NanoGPT and synthetic linear least squares even better than vanilla Muon.

@@

In this article, we tackle this question by actually showing that there are many viable non-operator norms. We leverage the family of norms dual to Ky Fan $k$-norms to derive a new family of Fanions, algorithms with low-rank updates. This approach theoretically explains the backbone of Dion's update \citep{ahn2025dion} and generalizes the memory-motivated application of the nuclear norm to Sharpness-Aware Minimization \citep{pethick2025sam}. As it was done with Muon, we come up with an effective procedure for computing Fanions' updates, Lanczos algorithm as described in ~\cref{sec:matrix-side} is the most effective algorithm, which, however, lacks an effective GPU- and PyTorch-friendly implementation.

>>>>

\author{Alexey Kravatskiy \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{kravtskii.aiu@phystech.edu} \\
\And
Ivan Kozyrev \\
Moscow Institute of Physics and Technology (MIPT), \\
Marchuk Institute of Numerical Mathematics \\
\texttt{kozyrev.in@phystech.edu} \\
\And
Nikolai Kozlov \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{kozlov.na@phystech.edu} \\
\And
Alexander Vinogradov \\
Moscow Institute of Physics and Technology (MIPT) \\
\texttt{vinogradov.am@phystech.edu} \\
\And
Daniil Merkulov \\
Moscow Institute of Physics and Technology (MIPT), Skoltech, HSE, AI4Science \\
\texttt{daniil.merkulov@phystech.edu} \\
\And
Ivan Oseledets \\
AIRI, Skoltech \\
\texttt{i.oseledets@skoltech.ru}
}

@@

\begin{abstract}
	In this article, we explore the use of various matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan $k$-norms to introduce a family of Muon-like algorithms we name {\it Fanions}, which happen to be similar to Dion. Then working with the duals of convex combinations of the Ky Fan $k$-norms and the Frobenius norm or the $l_\infty$ norm, we construct the families of {\it F-Fanions} and {\it S-Fanions} respectively. Their most prominent members are {\it F-Muon} and {\it S-Muon}. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that F-Muon and S-Muon are always on par with Muon and S-Muon, while on fine-tuning of NanoGPT and synthetic linear least squares even better than vanilla Muon.

@@

In this article, we tackle this question by actually showing that there are many viable non-operator norms. We leverage the family of norms dual to Ky Fan $k$-norms to derive a new family of Fanions, algorithms with low-rank updates. This approach theoretically explains the backbone of Dion's update \citep{ahn2025dion} and generalizes the memory-motivated application of the nuclear norm to Sharpness-Aware Minimization \citep{pethick2025sam}. As it was done with Muon, we come up with an effective procedure for computing Fanions' updates, Lanczos algorithm as described in~\cref{sec:matrix-side} is the most effective algorithm, which, however, lacks an effective GPU- and PyTorch-friendly implementation.
